---
title: "HW 3: Linear Regression Homework"
author: "Zhanchao Yang"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_float: 
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Introduction**
This assignment explores linear regression modeling using a new dataset, focusing on assumption checks and model interpretation. Students will fit a regression model, check assumptions, and interpret the results in a structured, easy-to-grade format.

---

## **1. Load Packages & Data**
```{r load_packages, message = FALSE}}
#install.packages(c("tidyverse", "broom", "ggplot2", "caret", "car", "MASS"))
library(tidyverse)
library(broom)
library(ggplot2)
library(caret)
library(car)
library(MASS)
```

### **Load New Dataset**
For this assignment, we will use the built-in `Boston` dataset from the `MASS` package, which contains housing data from different neighborhoods in Boston.
```{r load_data}
data("Boston")
boston_data <- as_tibble(Boston)
head(boston_data)
```

---

## **2. Exploratory Data Analysis**
### **2.1 Summary Statistics**
```{r summary_stats}
summary(boston_data)
```

### **2.2 Scatter Plots for Relationship Exploration**
```{r scatter_plots}
ggplot(boston_data, aes(x = rm, y = medv)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Relationship Between Number of Rooms and Median Home Value")
```

---

## **3. Fit a Linear Regression Model**
The response variable is `medv` (median home value), and predictors include `rm` (average rooms per house), `lstat` (percentage of lower-income population), and `crim` (crime rate per capita).
```{r linear_regression}
model <- lm(medv ~ rm + lstat + crim, data = boston_data)
summary(model)
```

### **Interpretation Questions:**
- What does the coefficient for `rm` suggest about the impact of more rooms on home prices?
- Does `lstat` negatively or positively affect home value? Is this expected?
- How well does the model fit the data (look at R² and Adjusted R²)?

---

## **4. Checking OLS Assumptions**
### **4.1 Residual Diagnostics**
```{r residual_diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(model)
```
- **Residuals vs Fitted Plot**: If a pattern exists, non-linearity may be present.
- **Q-Q Plot**: Checks if residuals are normally distributed.
- **Scale-Location Plot**: Detects heteroscedasticity.
- **Residuals vs Leverage**: Identifies influential observations.

### **4.2 Multicollinearity Check**
```{r vif_check}
vif(model)
```
- VIF values above **5** indicate high multicollinearity.

---

## **5. Model Refinement: Log Transformations**
If assumptions are violated, log-transforming variables can improve the model.
```{r log_transform}
boston_data <- boston_data %>% mutate(
  log_medv = log(medv),
  log_lstat = log(lstat + 1)
)

log_model <- lm(log_medv ~ rm + log_lstat + crim, data = boston_data)
summary(log_model)
```
- Compare the R² of both models. Does the transformation improve fit?

---

## **6. Model Performance & Cross-Validation**
```{r cross_validation}
set.seed(123)
train_control <- trainControl(method = "cv", number = 10)
cv_model <- train(medv ~ rm + lstat + crim, data = boston_data, method = "lm", trControl = train_control)
cv_model
```

---

## **7. Homework Assignment (10 Points)**
Answer the following questions and submit your answer in a nice R Markdown file.

### **Part 1: Model Interpretation (4 Points)**
1. What does the coefficient for `rm` mean in the original model? (1 point)
Answer: 
2. How does `lstat` impact median home value, and why? (1 point)

3. Is crime rate (`crim`) statistically significant? Justify using the p-value. (1 point)

4. How well does the original model explain home values (interpret R² and Adjusted R²)? (1 point)

### **Part 2: Assumption Checks & Model Improvement (4 Points)**
5. Based on residual plots, is the model violating any OLS assumptions? (1 point)
6. What does the VIF test indicate about multicollinearity? (1 point)
7. After log-transforming `lstat`, does model performance improve? Explain. (1 point)
8. Compare RMSE from cross-validation to the model's residual standard error. Which suggests better predictive performance? (1 point)

### **Part 3: Expanding the Model (2 Points)**
9. Add `dis` (distance to employment centers) to the model. Does it improve fit? (1 point)
10. Try another predictor from the dataset that you think might be relevant. Justify why you selected it and interpret its impact on the model. (1 point)
11. Wasn't this fun? (0 Points)



## **Submission Instructions**
- Submit a **fully documented R script or R Markdown file**.
- Include **plots and model summaries** where relevant.
- Provide **concise interpretations** for all responses.

---